---
title: "Calculation of Age- and Sex-Specific Cut-Points"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculation of Age- and Sex-Specific Cut-Points}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
csl: biomed-central.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The R package `popmetrics` facilitates the calculation of age- and
sex-specific cut-points for physical activity intensity classification
based on the free-living accelerometer data recordings of your study
sample. This methodology has been tested for first time using the NHANES
data, specifically the first wave of accelerometer data, including the
2003-2004, and 2005-2006 cohorts @Migueles2024.

No criterion measure is needed to extract the age- and sex-specific
cut-points as they are based on the population (or study sample)
descriptive values and their recorded free-living physical activity
patterns.

## Input

The expected input data that you need to get the age- and sex-specific
cut-points consists of two main pieces. First, data frame(s) with
epoch-level aggregation of acceleration values, and second, the
descriptive values to be considered when fitting the cut-points to the
population (e.g., age and sex for each participant in the data set).

#### Acceleration values

There are two possibilities, either providing a single data frame
containing the information for all the participants in the study sample,
or providing data frames for each individual.

-   In the first case, the user should provide the data frame as an R
    object of class data.frame or data.table containing three columns:
    ID, time stamp, and acceleration metric.

-   In the second scenario, the user can provide the path to a folder
    containing csv or RData files for each individual. In this case,
    `popmetrics` would derive the ID from the file names by eliminating
    the file extension (e.g., `id001.csv` would result in `id001`).
    Thus, the data frames should only contain two columns with the time
    stamp and the acceleration metric.

The format of the columns is expected to be as follows:

-   **ID column:** it can contain either character or numeric
    specification of the participants' IDs.

-   **Time stamp column:** it should contain the beginning of each epoch
    either as a `POSIX`, `ISO8601`, or a `character` with format
    specified by the user (following the standards explained in the
    documentation of `base::strptime`.

-   **Acceleration values:** it should contain the acceleration metric
    that would be used in the calculation of the cut-points.

#### Descriptive values

A second data frame is expected from the user containing the descriptive
values of the study sample. This data frame should contain a column with
the ID, and as many columns as factors should be used in the calculation
(e.g., ID, sex, age). It is imperative that the ID in this data frame
exactly matches the ID extracted or specified in the acceleration
values.

A specific terminology is used for the specification of the factors in
the `popmetrics` functions:

-   `factor_split` refers to any factor that would be used to split the
    sample before fitting the distribution of the accelerations.
    Usually, this would be a factor with not many levels (e.g., sex).

-   `factor_fit` refers to the factor over which the GAMLSS model would
    estimate the distribution of the acceleration metric. Usually, this
    includes a finer gradient of levels (e.g., age).

## Methodology

The methodology for calculating the age- and sex-specific cut-points for
moderate-to-vigorous physical activity are detailed in the study by
Migueles et al. @Migueles2024, with a deeper explanation in the
Supplemental file.

In brief, the methodology consists of the following steps:

1.  Calculation of the maximum 5-min intensity for each day of the
    free-living recording, inspired in the methods proposed by Rowlands
    et al. @Rowlands2018.

2.  Extraction of the maximum value of the daily maximum 5-min intensity
    for each participant.

3.  Fitting of the maximum 5-min intensity for each participant across
    the sexes and ages using GAMLSS modelling.

4.  Definition of the target intensity as the percentile 80 of the
    maximum 5-min intensity for each age and sex.

5.  Definition of the cut-point for moderate-to-vigorous physical
    activity as the 40% of the target intensity for each age and sex.

## Example use-case: NHANES '03-'04 cohort

Hereby there is an example for calculation of the age and sex-specific
cut-points as it was conducted in the proof of concept publication
@Migueles2024. Loading the package allow access to the data and a
function to facilitate some calculations.

```{r setup}
library(popmetrics)
```

### 1. Read data in

In this tutorial, we will be using the NHANES data from 2003-2004, and
2005-2006. The data comes in the format of `.xpt` containing the data
for all participants at minute level. Data is publicly available at
<https://wwwn.cdc.gov/Nchs/Nhanes/>

```{r eval=FALSE}

# ----------------------------
# DOWNLOAD ACCELEROMETER DATA FROM NHANES 2003-2004:
# The NHANES files are ~400 MB. 
# You may need to extend the time for the download, you can do that with: 
# The download and reading may take some time, be patient =)
# ----------------------------
options(timeout = 60*20)
temp = tempfile(fileext = ".zip")
download.file(url = "https://wwwn.cdc.gov/Nchs/Nhanes/2003-2004/PAXRAW_C.ZIP",
              destfile = temp, mode = "wb")
acc_min = foreign::read.xport(unzip(zipfile = temp, files = "paxraw_c.xpt"))
unlink(temp)

# ----------------------------
# DOWNLOAD DESCRIPTIVE DATA FROM NHANES 2003-2004:
# ----------------------------
temp = tempfile(fileext = ".XPT")
download.file(url = "https://wwwn.cdc.gov/Nchs/Nhanes/2003-2004/DEMO_C.XPT",
              destfile = temp, mode = "wb")
desc = foreign::read.xport(temp)
unlink(temp)

# ----------------------------
# Note that in the published study we use data from 2003-2004, and 2005-2006.
# To replicate that, download both files and rbind them.
# ----------------------------

```

Now, `acc_min` is a data frame containing time stamp (columns `PAXHOUR`
and `PAXMINUT`) and the acceleration metric (column `PAXINTEN`). In this
case, the acceleration metric is **counts per minute** as derived from
the **ActiGraph 7164** at that time.

### 2. Obtaining the age- and sex-specific cut-points

#### Step 1 - get the maximum 5-min intensity per day

```{r eval=FALSE}

# Extract maximum 5-min intensity per day
maxDay = aggregate(PAXINTEN ~ PAXDAY + SEQN, data = acc_min, FUN = maxth,
                       k = 5, epoch = 60, consecutive = FALSE)
rm(acc_min) # no longer needed

```

#### Step 2 - clean data set

Prior to getting the maximum value of the recording per participant, we
need to make sure that we only fit days with valid, reliable, and
representative data of each day. For example, we need to remove files
without enough valid data in the recording, or files that were not
correctly calibrated. In the example of the NHANES data:

```{r eval=FALSE}

# ----------------------------
# Wear time indicator per day for this dataset is available as popmetrics::day_wear
# ----------------------------

# ----------------------------
# Keep files with:
# PAXCAL = 1 & PAXSTAT = 1 & PAXWEARMINUT >= 600 (10h of wear time)
# ----------------------------
include = which(day_wear$PAXCAL == 1 & 
                  day_wear$PAXSTAT == 1 & 
                  day_wear$PAXWEARMINUT >= 600)
days2include = paste(day_wear[include,"SEQN"], day_wear[include,"PAXDAY"])
dayIndex = paste(maxDay[,"SEQN"], maxDay[,"PAXDAY"])
maxDay_clean = maxDay[which(dayIndex %in% days2include),]
```

#### Step 3 - get the maximum 5-min intensity per each participant

Maximum observed over the days in the recording.

```{r eval=FALSE}

# Keep maximum max 5-min intensity of the recording per participant
maxWeek = aggregate(PAXINTEN ~ SEQN, data = maxDay_clean, FUN = max)

```

### Step 4 - fit distribution over the specified factors

`factor_split` (sex) and `factor_along` (age)

```{r eval=FALSE}

library(gamlss)

# data to fit (ID, age, sex, max5intensity)
data2fit = merge(desc[, c("SEQN", "RIAGENDR", "RIDAGEYR")], maxWeek, by = "SEQN")

# GAMLSS fitted distribution for Men
distribution_men = gamlss(PAXINTEN ~ cs(RIDAGEYR, df = 2),
                          sigma.formula = ~cs(RIDAGEYR, df = 2),
                          nu.formula = ~cs(RIDAGEYR,df=2), 
                          family = BCCG,
                          data = subset(data2fit, RIAGENDR == 1))

# GAMLSS fitted distribution for Men
distribution_women = gamlss(PAXINTEN ~ cs(RIDAGEYR, df = 2),
                            sigma.formula = ~cs(RIDAGEYR, df = 2),
                            nu.formula = ~cs(RIDAGEYR,df=2), 
                            family = BCCG,
                            data = subset(data2fit, RIAGENDR == 2))

```

#### Step 4 - Define target intensity and cut-points

Percentile 80 of the distribution and 40% of that

```{r eval=FALSE}

# men
target_men = centiles.pred(distribution_men,
                           type = "centiles", 
                           xname = "RIDAGEYR", 
                           xvalues =  subset(data2fit, RIAGENDR == 1)$RIDAGEYR, 
                           yval = subset(data2fit, RIAGENDR == 1)$PAXINTEN,
                           cent = 80)
years = c(1, which(diff(target_men$x) == 1) + 1)
target_men = cbind(target_men[years,1], 
                    round(target_men[years,2:ncol(target_men)]))
target_men$MVPA_threshold = target_men[, 2] * 0.4


# women
target_women = centiles.pred(distribution_women,
                             type = "centiles", 
                             xname = "RIDAGEYR", 
                             xvalues =  subset(data2fit, RIAGENDR == 2)$RIDAGEYR, 
                             yval = subset(data2fit, RIAGENDR == 2)$PAXINTEN,
                             cent = 80)
years = c(1, which(diff(target_women$x) == 1) + 1)
target_women = cbind(target_women[years,1], 
                    round(target_women[years,2:ncol(target_women)]))

target_women$MVPA_threshold = target_women[, 2] * 0.4


```

## Future potential applications

This methodology was specifically developed for physical activity
surveillance, yet at the moment, only one study testing this as a proof
of concept has been published @Migueles2024. This R package has been
built to facilitate further development, which will include the
following.

### Other acceleration signal aggregation metrics

Preferably, the acceleration aggregation measure would be calculated
following any of the methodologies proposed in the physical activity
literature. Examples include (but are not limited to) ENMO @van2013, MAD
@Aittasalo2015, MIMS @John2019, AI0 @Bai2016, or activity counts
@Neishabouri2022.

### Other data collection protocols

As this methodology is specific to the acceleration patterns recorded,
we hypothesize that it will work with study protocols including other
acceleration brands, body attachment sites, sampling frequencies, among
others. The only important requirement is that the data is
representative of the usual lifestyle of the population, including at
least 7-day long recordings.

### Using other population factors to develop the cut-points

In the proof of concept, we used age and sex to make the cut-points
specific to the population characteristics. Age and sex may be enough
for the general population, yet this approach can also be adapted to
specific factors such as socioeconomic status, weight status levels,
physical fitness, or presence of any pathology, among others.

## References
